```{r results='hide', message=FALSE, warning=FALSE}
library(caret)
```
Practical Machine Learning Course Project
========================================================
## Executive Summary
This project explores Human Activity Recognition. Specifically it
predicts the manner in which an exercise was performed.
The project data is from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

The data is already separated into training and testing sets:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

In order to determine the optimal parameters the training data will be further
separated into training and validation subsets.

The data has been downloaded into the current directory in order to run the prediction.

## Exploratory Data Analysis
```{r}
trainingData <- read.csv('pml-training.csv')
testingData <- read.csv('pml-testing.csv')
```
The training dataset is composed of `r dim(trainingData)[1]` observations of `r dim(trainingData)[2]` variables.
The testing dataset is composed of `r dim(testingData)[1]` observations of `r dim(testingData)[2]` variables.
The variables are all interpeted as numeric although several are ordinal.
The dependent (output) variable in the training data is **classe**, the method of exercise performed.
In the testing dataset this variable is replaced by **problem_id** used in the project submission.

The first 7 variables are not sensor readings but timing and user information.
These are not needed for the analysis and are removed.
All columns consisting of NA data are also removed.

```{r}
trainingData <- trainingData[8:160]
testingData <- testingData[8:160]
testingData <- testingData[,(colSums(is.na(trainingData)) == 0)]
trainingData <- trainingData[,(colSums(is.na(trainingData)) == 0)]
```
The resulting data still has `r dim(trainingData)[2]` variables.
We will remove near zero variance variables which would not influence the outcome and highly correlated variables which are redundant.
```{r}
nsv<- nearZeroVar( trainingData, saveMetrics=TRUE)
list <- labels(nsv[ nsv$nzv == TRUE,])[1]
l <- unlist(list)
trainingData <- trainingData[, -which(names(trainingData) %in% l)]
testingData <- testingData[, -which(names(testingData) %in% l)]
```
The resulting data still has `r dim(trainingData)[2]` variables.
Now we will remoe highly correlated variables.
```{r}
M <- abs( cor( trainingData[,-dim(trainingData)[2]]))
corrValues <- findCorrelation( M, cutoff = .80, verbose = FALSE)
trainingData <- trainingData[, -corrValues]
testingData <- testingData[, -corrValues]
```
Now we will split the data into training and validation sets.
```{r}
inTrain <- createDataPartition( y = trainingData$classe, p=.5, list=FALSE)
training <- trainingData[inTrain,]
cv <- trainingData[-inTrain,]
```
Note that this was performed on a Windows machine and uses the doSnow library to
make use of multiple cores. Update accordingly for different platforms.

```{r  results='hide', message=FALSE, warning=FALSE}
library("doSNOW")
```
```{r cache=TRUE}
cl<-makeCluster(4)
registerDoSNOW(cl) # Register the cores.
model <- train( classe~., data = training, method="rf", prox=TRUE)
stopCluster(cl)

model
```
The model has excellent results, with an in sample error rate of 1.6%.
We would expect the out of sample error to be somewhat higher and will estimate it using
the cross validation data.

```{r}
cvPredict <- predict( model, cv)
confusionMatrix( cvPredict, cv$classe)
```
The out of sample error estimate is .5%, so we do not need to do any more modification of the training model.
Now we perform prediction on the test data and produce test files for submission.

```{r}
prediction <- predict( model, testingData)
write_files <- function(x) {
        n <- length(x)
        for (i in 1:n) {
                filename <- paste0("problem_id", i, ".txt")
                write.table(x[i], file=filename, quote=FALSE, row.names=FALSE,col.names=FALSE)
        }
}
write_files(prediction)
```